Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network

### 2021-01-21 docs/manuscript/manuscript.tex

\section{Introduction}
Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been
done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions). In general,
tree-related outages are due to contact between tree parts and power lines. Sometimes, the contact is due to growth of
branches into the lines; sometimes it is due to branch failure; sometimes it is due to tree failure (either uprooting or
trunk failure). When outages occur, the economic costs can be substantial. For example, from 2005 to 2015,
\cite{graziano2020wider} computed the annual economic cost of outages in the US state of Connecticut; the mean was
approximately \$8.3 bn.  Outages due to contact from branch growth or branch failure can be mitigated by line clearance
tree trimming. Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in
the billions of dollars \cite{guggenmoos2003effects}. Despite the comparatively high annual cost of tree trimming,
studies have repeatedly shown that trimming branches away from the wires reduces the number of outages (sometimes
classified as “preventable” outages). In Massachusetts, USA, for instance, tree failure was responsible 40\% of preventable
tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\%
\cite{simpson1996treecaused}. A more recent (and sophisticated) analysis in Connecticut showed that enhanced tree
trimming reduced outages during storms \cite{parent2019analysis}. And using a statistical model to predict outages based
on data from two Gulf Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less
uncertain when they included the effect of tree trimming; models including wind speed without considering the effect of
trimming were not as accurate.

Trimming can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging
the lines in proximity are removed, reducing the likelihood of contact.
% Trimming also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of
% whole-tree failure, whether by uprooting or stem breakage.  Although the annual cost of trimming is comparatively
% large, the economic cost of outages can far exceed it. A recent benefit cost analysis demonstrated a highly favorable
% benefit cost ratio for trimming relative to the economic costs of outages (Graziano et al. 2020).  Trimming,
However, it cannot eliminate tree-related outages. Failure of trees away from the right-of-way can still impact the
lines and cause outages \cite{guggenmoos2003effects}. The proportion of tree failures away from the wires that cause
outages varies and has not been rigorously quantified. \citet{guggenmoos2011treerelated} estimated that 95\% of
tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted}
reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been
developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Risk includes assessing the likelihood of
failure, the likelihood of impact, and the severity of consequences. The likelihood of failure depends on the anticipated
loads on the tree and its load-bearing capacity. The likelihood of impact depends on proximity to the target (the lines,
poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate
(which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees. Severity of
consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs
and disruption associated with outages---this, in turn, is partially related to the size of the tree or tree part that
fails, and how much momentum it has when it impacts the infrastructure.

@@

\section{Data and Methods}
\subsection{Image data description}
The training dataset consists of 505 images, each having an original size of $4032\times3024$ pixels. 
Based on the  ISA’s Tree Risk Assessment Qualification (TRAQ) protocol, four categories of tree failure likelihood are defined. In this paper we only focus on three:
\begin{itemize}
\item \textbf{Probable}: failure expected under normal weather conditions within a given timeframe
\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
\end{itemize}

>>>>

\section{Introduction}
Utilities have been concerned with tree-related outages for many years, but much of the work to quantify them has been
done in-house: Utilities contract with a consultant who analyzes tree-related outages (or interruptions). In general,
tree-related outages are due to contact between tree parts and power lines. Sometimes, the contact is due to growth of
branches into the lines; sometimes it is due to branch failure; sometimes it is due to whole-tree failure (either uprooting or
trunk failure). When outages occur, the economic costs can be substantial. For example, from 2005 to 2015,
\cite{graziano2020wider} computed the annual economic cost of outages in the US state of Connecticut; the mean was
approximately \$8.3 bn.  Outages due to contact from branch growth or branch failure can be mitigated by line clearance
pruning. Estimates vary, but the annual amount that utilities spend on tree trimming operations is typically in
the billions of dollars \cite{guggenmoos2003effects}. Despite the comparatively high annual cost of pruning,
studies have repeatedly shown that pruning branches away from the wires reduces the number of outages (sometimes
classified as “preventable” outages). In Massachusetts, USA, for instance, tree failure was responsible 40\% of preventable
tree-caused outages, but removing or severely pruning high risk trees improved reliability by 20\% to 30\%
\cite{simpson1996treecaused}. A more recent (and sophisticated) analysis in Connecticut showed that enhanced tree
pruning reduced outages during storms \cite{parent2019analysis}. And using a statistical model to predict outages based
on data from two Gulf Coast states in the USA, \citet{nateghi2014power} demonstrated that prediction models were less
uncertain when they included the effect of utility pruning; models including wind speed without considering the effect of
trimming were not as accurate.

%Utility pruning can be effective because both structurally-deficient and structurally-sound branches adjacent to or overhanging
%the lines in proximity are removed or reduced, reducing the likelihood of contact.
Utility pruning aims to reduce the frequency of tree-caused outages through the reduction or removal of branches in proximity to overhead electrical lines.
% Pruning also reduces external loads on the tree (e.g., wind or snow and ice), which reduces the likelihood of
% whole-tree failure, whether by uprooting or stem breakage.  Although the annual cost of utility vegetation management is 
% large, the economic cost of outages can far exceed it. A recent benefit cost analysis demonstrated a highly favorable
% benefit cost ratio for pruning relative to the economic costs of outages (Graziano et al. 2020).  
However, it cannot completely eliminate tree-related outages. Failure of trees away from the right-of-way can still impact the
lines and cause outages \cite{guggenmoos2003effects}. The proportion of tree failures away from the wires that cause
outages varies and has not been rigorously quantified. \citet{guggenmoos2011treerelated} estimated that 95\% of
tree-caused outages in the Pacific Northwest region of the US, were due to tree failure, and \citet{wismer2018targeted}
reported approximately 25\% of interruptions in Illinois, USA, were caused by trees that uprooted or broke in the stem.
Predicting the likelihood of failure is an inexact science, but tree risk assessment best management practices have been
developed \cite{e.thomassmiley2017best,johnw.goodfellow2020best}. Risk includes assessing the likelihood of
failure, the likelihood of impact, and the severity of consequences. The likelihood of failure depends on the anticipated
loads on the tree and its load-bearing capacity. The likelihood of impact depends on proximity to the target (the lines,
poles, and other hardware---``infrastructure''---in the case of utility tree risk assessment), the target’s occupancy rate
(which is constant for utility lines) and whether the target is sheltered, for example by neighboring trees. Severity of
consequences depends on the damage done to the infrastructure and, more importantly in some cases, the economic costs
and disruption associated with outages---this, in turn, is partially related to the size of the tree or tree part that
fails, and how much momentum it has when it impacts the infrastructure.

@@

\section{Data and Methods}
\subsection{Image data description}
The training dataset consists of 505 images, each having an original size of $4032\times3024$ pixels. 
Based on the  ISA’s Tree Risk Assessment protocol, four categories of tree failure likelihood are defined. In this paper we only focus on three:
\begin{itemize}
\item \textbf{Probable}: failure expected under normal weather conditions within a given timeframe
\item \textbf{Possible}: failure expected under extreme weather conditions; but unlikely during normal weather conditions
\item \textbf{Improbable}: failure unlikely either during normal or extreme weather conditions 
\end{itemize}
